Supplementary materials for 'Canaries of the Arctic: the collapse of eastern Bering Sea snow crab

### 2023-04-13 manuscript_supmat.Rmd
revisions commit

---
author: ""
date:
output:
  pdf_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    keep_tex: true
    includes:
      in_header: my_header.tex
header-includes: 
  - \usepackage[left]{lineno}
  - \linenumbers
#-  \pagenumbering{gobble}
csl: fish-and-fisheries.csl
title: "Supplementary materials: Canaries of the Arctic: the collapse of eastern Bering Sea snow crab"
---

@@

# Supplementary materials
## Methods overview
We used an integrated population model to estimate variation in mortality over time for snow crab in the eastern Bering Sea and generalized additive models (GAMs) to relate the estimated variation in mortality to potential stressors in the environment.  The population dynamics model was fit to abundance and size composition data from the National Marine Fisheries Service (NMFS) summer bottom trawl survey in the eastern Bering Sea shelf to estimate total mortality by maturity state and year for male snow crab. We then developed indices for temperature occupied, disease prevalence, cannibalism, and crab density from the NMFS survey to test as covariates in GAMs. Indices for fishery related effects were collated from fisheries statistics from the Alaska Department of Fish and Wildlife and also included in the GAMs.  Below we describe each of these components, discuss the rationale behind our modeling decisions, and provide sensitivities and simulation tests of our models.

## Population dynamics model
The population dynamics model presented here incorporates the best available information on relevant population processes to estimate total mortality for male snow crab on the eastern Bering Sea shelf and is similar in structure to the model used to assess eastern Bering Sea snow crab for management (Szuwalski, 2021). The model tracked numbers of male crab at size at maturity state over time with size bins ranging from 30-95 mm carapace width with 5 mm bin widths. Only male crab were modeled because male and female crab appear to have somewhat different dynamics and the male crab in the modeled size range are better selected by the survey gear (Szuwalski, 2021). Snow crab are sexually dimorphic, with male snow crab growing to nearly twice the size of females, which accounts for the better selection in the survey. Only crab smaller than 95 mm were modeled for two reasons: 1) to attempt to isolate the effect of the directed fishery (crabs of >101 mm carapace width are targeted in the fishery; discussed further below) and 2) almost all of the crab that disappeared since 2018 are in this size range. The population dynamics model operates on a half year time step, starting in July at the time of the NMFS survey. The fishery is assumed to occur in February. Total mortality (Z) is estimated by year (y) and maturity state (m). Other estimated parameters include the initial numbers at size by maturity state, yearly log recruitments, a vector of scalars that determine the proportions of estimated recruitment split into the first two size bins, and a variance component for the penalty on total mortality. Parameters determining growth, maturity, and survey selectivity were estimated outside of the model and specified when estimating mortality and catchability. Mortality is the only population process that occurs in the first half of a given year:

@@

The size transition matrix $X_{s,s'}$ was constructed using growth increment data collected over several years (see Szuwalski, 2021 for a summary) to estimate a linear relationship between pre- and post-molt carapace width (\autoref{linear_growth}), ($\hat{W}^{pre}_{s,w}$ and $\hat{W}^{post}_{s,w}$, respectively) and the variability around that relationship was characterized by a discretized and renormalized normal distribution with a size-varying standard deviation, Y~*s,w,w'*~ (\autoref{linear_growth}).

@@

Finally, the last half of the year of mortality is applied to the population after growth, maturity, and recruitment occurs. Note that this allows a crab to experience two different mortalities within a given year as it undergoes terminal molt.

@@

The observed numbers of crab at size by year in the NMFS survey reflect the ability of the trawl gear to capture the crab, also known as 'selectivity'. The selectivity of trawl gear can change according to size, and consequently needs to be accounted for in the population dynamics model when fitting to the survey data. Values for survey selectivity at size were specified using data from experimental trawls conducted by the Bering Sea Fisheries Research Foundation in collaboration with the NMFS summer survey. The experimental trawls were performed at the same time and location as the NMFS summer survey tows to evaluate the efficiency of the NMFS survey trawl gear at capturing snow crab (Somerton et al., 2013). The nephrops gear used by the BSFRF was assumed to capture all crab in its path given strong bottom contact.  The resulting area-swept estimates of numbers of crab at size from the BSFRF and NMFS surveys ($\hat{N}_{y,s,NMFS}$ and $\hat{N}_{y,s,NMFS}$, respectively) can be used to infer the selectivity of the NMFS gear in year y as:

 \begin{equation} S_{y,NMFS} = \frac {\hat{N_{y,s,NMFS}}}{\hat{N}_{y,s,BSFRF}} \end{equation}

@@

## Objective function
The objective function for the population dynamics model consists of likelihood components and penalty components that are summed and minimized in log space to estimate parameters within the model. Several data sources were fit to using the following likelihoods. Observed size composition data for immature and mature males were fit using multinomial likelihoods and were implemented in the form:

@@

L~*x*~ was the likelihood associated with data component x, where $\lambda_{x}$ represented an optional additional weighting factor for the likelihood, $N_{x,y}$ was the sample sizes for the likelihood, $p^{obs}_{x,y,l}$ was the observed proportion in size bin *l* during year *y* for data component *x*, and $\hat {p}_{x,y,l}$ was the predicted proportion in size bin *l* during year *y* for data component *x*. Sample sizes were input as 100, which is the value currently used in the stock assessment.

@@

Smoothing penalties were placed on estimated vectors of deviations for immature and mature natural mortality and immature and mature catchability. A prior value of 0.27 is used for the average natural mortality based on assumed maximum age of 20 and Hamel's (2015) empirical analysis of life history correlates with natural mortality. The priors used for catchability were derived from the selectivity experiments described above. Penalties were implemented using normal likelihoods on the second differences of the vector. A separate normal prior was placed on the estimated mean value of immature and mature mortality in the form:

\begin{equation} L_{x} = \lambda_{x} \sum_{y} \frac {((\hat{I}_{x,y}) - (I_{x,y}))^2}{CV_{x,y}^2}
\end{equation}

## Population dynamics model sensitivities
Modeling decisions are necessarily made in the process of writing population dynamics models and it is possible for these decisions to influence the outcome of an analysis. Within the context of our model, these decisions include what processes to allow to vary over time, the weights assigned to different data sources and portions of the objective function, which parameters to place priors or penalties on, and what those priors or penalties should be. We ran several sensitivity analyses to understand the implications of these modeling decisions on the outcome of our analysis.

### Does allowing mortality or catchability to vary over time improve model fits?
Catchability and mortality are somewhat confounded within population dynamics models (Thompson, 1994). Fewer crab observed in a given year can be attributed to either crab dying or by crab moving out of the surveyed area either by walking out of the boundaries or burying themselves into the substrate. At the same time, it is also clear that catchability and mortality likely vary over time in reality in spite of the fact that they are often assumed to be time-invariant in population dynamics models (Johnson et al., 2014). Somerton et al. (2013) showed that catchability varied by substrate and depth for snow crab in the EBS. The spatial distribution of snow crab varies over time and substrate and depth vary over space, so it follows that catchability should also vary over time.

We started exploring the impacts of including time-variation in mortality and catchability on model output by  fitting a model with no time-variation in mortality or catchability. Then we compared the output of this model to models that allow time-variation in mortality, catchability, and both processes simultaneously (\autoref{inc_complex} & \autoref{inc_comp_proc}). The model with no time-variation in mortality or catchability was able to capture the general trend in immature and mature survey abundance solely through estimating variability in recruitment. Allowing time-variation in catchability improved the fits to immature survey abundances more than time-varying mortality, but time-variation in either process improved fits in a similar manner for mature survey abundances. Mature size composition data were fit similarly for all models, but immature size composition data were better fit by the models that allowed time-varying catchability (\autoref{inc_complex}). Part of the reason this difference in fits to immature size composition data occurs is the variability in the first several size bins resulting from the poor selectivity of the survey for small animals. Sometimes the peaks seen in larger size classes are reflected in the preceding years' data for the smallest size classes, sometimes those peaks are not reflected (compare \autoref{expand_size_fits_imm} to \autoref{expand_size_fits_mat}). As a consequence, positive residuals occur in the smallest size classes when a pseudocohort is consistently seen in large size classes, but not observed in the smallest size bins (e.g. 1991 vs. 1992; 1997 vs 1998).

@@

Differences among sensitivity scenarios resulted in very small changes in the fits to the data (\autoref{sens_fits}), but larger changes in estimated mortalities and catchabilities (\autoref{sens_ests}). The smoothness penalty placed on mortality over time appeared to be the largest driver of changes in estimates of M and q, so we looked at a wider range of smoothness penalties (i.e. 0.001, 0.1, 0.25, 0.5, 1, 5, 10, 1000).  Trajectories of mortalities were roughly preserved across this range. The prior on mean natural mortality predictably scaled the estimated time series up or down. The best available information suggests natural mortality should be approximately 0.27 given an assumed (but based on a range of studies; see Szuwalski, 2021 for a summary) maximum age of 20 years for wild snow crab. Based on these analyses, we elected to use small smoothing penalties because there is no evidence to suggest that mortality should be particularly smooth from year to year and relatively tight priors on the mean mortality given outside information to support this assumption. These analyses also underscore the fact that the scale of the population is difficult to estimate with the data available and the need to relate mortality to the environmental covariates outside of the population dynamics model. This likely comes from the fact that recruitment and immature mortality are confounded (i.e. fewer immature crab in a given year can be because of increased immature mortality or because of lower recruitment).

@@

Another important point for consideration in covariate construction is the estimation of mortality by maturity state. Snow crab in the EBS undergo an ontogenetic migration in which juvenile crab settle on the northeast portion of the shelf after their pelagic phase, then migrate southwest into deeper and (usually) warmer waters (Ernst et al., 2005; Parada et al., 2010). This means that the conditions and stressors experienced by immature crab can be different than those by mature crab. To address this issue, the spatial data sets for temperature, disease, and cannibalism were split based on the size above which half of the population was mature. The size at which more than half of the population is mature changes by year, depending on recruitment dynamics and other demographic processes (\autoref{split_mat}). After the survey data were split at the 50% at maturity size, time series of maturity-specific environmental stressors (\autoref{grid_covar}) were created as described below.

### Temperature
Temperature is one of the key physical variables that structures the benthic ecosystem of the EBS (Mueter and Litzow, 2008). The cold pool, a mass of water <2 degrees Celsius, acts as a barrier to species interaction based on temperature preferences of different species. Snow crab are a stenothermic species, preferring cold water and juvenile snow crab in particular are rarely found outside of the cold pool (Dionne, 2003). The cold pool is directly related to the winter ice extent in the Bering Sea and has varied dramatically over time as the ecosystem moves between cool and warm stanzas (e.g. 2006-2010 vs. 2014-2019; Figure 1b of the main text and \autoref{bottom_temp}). As the cold pool changes from year to year, so does the spatial distribution of snow crab (\autoref{snow_dist}). The ontogenetic migration of snow crab results in crab of different sizes and maturity states experiencing different temperatures in a given year (\autoref{temp_occupied}). The 'temperature occupied' for different sizes of crab by year $T_{s,y}$ was calculated here as an average of the observed bottom temperatures at the stations at which crab of a given size were captured $t_{i}$, weighted by the area-swept density of crab at a given station $d_{i}$:

@@

### Predation
Pacific cod (*Gadus macrocephalus*) are the most important predator of snow crab based on stomach content data collected in the NMFS bottom trawl survey (Long and Livingston, 1998). Immature crab under the size of 55 mm carapace width are the primary sizes consumed by cod in the Bering Sea (Burgos et al., 2010). An index of summer daily consumption (tons/day) of snow crab between 30-95mm carapace width eaten by Pacific cod in the eastern Bering Sea was developed using cod stomach content data from the survey to estimate the proportion by weight of crab in cod diets and the size composition of crab by carapace width of prey found in cod stomachs, stratified by year, survey stratum, and cod length (collection and analysis methods described in Livingston et al. 2017).  Cod total consumption rate (metabolic demand) was calculated using a cod bioenergetics model (Holsman and Aydin 2015) to estimate laboratory-measured maximum consumption rates adjusted for bottom water temperatures and cod abundance-at-length measured at each haul location (following methods described in Barbeaux et al. 2020), and summed to an ecosystem-wide total.

Changes in the cold pool can alter the interaction between snow crab and Pacific cod over time. Decreases in the size of the cold pool coincide with more northerly positions of the centroids of abundance of cod (e.g. 2003 and 2018-2019; \autoref{cod_centroids} & \autoref{cod_pred_data}).  This increased interaction coincided with increased numbers of crab consumed by cod in the last several years (\autoref{cod_consumption_raw}). However, this period of time also coincided with the appearance of the largest pseudo-cohort of snow crab ever seen in the Bering Sea. Given the generalist nature of Pacific cod, one would expect to see an increase in the amount of crab consumed by cod during this period of time even if there weren't differences in the interactions between the species as a result of changes in the cold pool. Further, a large fraction of the missing crab from the recent collapse were not of the sizes typically eaten by cod (\autoref{sizes_of_missing_crab}). To evaluate the possibility cod consumption has influenced the mortality of snow crab over time, the relative impact of consumption with respect to the population size must be considered. The vast majority of snow crab consumed by Pacific cod are immature, so only indices of cod predation for immature crab were included in the models below. Predation indices were calculated for immature animals by year $P_{m,y}$ by calculating the ratio of the extrapolated biomass of crab consumed by cod to the biomass of the estimated biomass of crab, $N_{y,m,s}*w_{s}$:

@@

The exact amount of crab eaten cannot be calculated from the available diet data because they are a snapshot of consumption at one point during the year and consumption would be expected to change with spatial overlap and temperature-driven changes in metabolism occurring throughout the year. Consequently, removals due to predation cannot be directly incorporated into the model as fishery removals might be. However, the predation index developed here represents the best available information on the relative impact of cod predation on snow crab mortality for use in correlative models like the GAMs below.

@@

The proportion of smaller than 55 mm carapace width crab overlapping with larger than 95 mm carapace width crab was calculated by finding the intersection of the station IDs at which at least one crab of both size classes was observed. The density of crab larger than 95 mm carapace width was calculated as the number of >95 mm carapace width crab observed at those stations multiplied by the area swept. This exercise was also done by 5 mm size bins to show the overlap of small crab of different sizes with large crab (\autoref{cannibalism_size_raw}). The final index aggregated all crab smaller than 55 mm carapace width (\autoref{cannibalism}). Indices of cannibalism were only included in the immature models given laboratory observations indicate cannibalism is rare among crab of similar sizes.

### Fisheries data
Snow crab are caught both in a directed fishery (i.e. a fishery aimed at capturing snow crab) and non-directed fisheries (i.e. fisheries with targets other than snow crab). In the directed fishery, under-sized and/or dirty shelled crab are often discarded. Snow crab are discarded from non-directed fisheries using a variety of gear types (including trawl, pots, hook-and-line) and targeting a variety of species (e.g. Pacific cod, walleye pollock, and yellowfin sole) that operate over a wide fraction of the Bering Sea shelf (\autoref{bycatch_binned}). \autoref{bycatch_binned} is plotted in log space, so it appears that the bycatch is spread widely over the shelf, but in normal space, the bycatch is more concentrated (e.g. \autoref{bycatch_binned_2018}). The location of the centroids of the bycatch have moved over time and increases in latitude correspond with warm years in which reduced ice extent allowed for fishing farther north (\autoref{bycatch_centroids}). Bycatch in groundfish trawl fisheries are by far the largest sources of bycatch mortality (\autoref{bycatch}). Data on discards and bycatch of snow crab are collected by at-sea observers on fishing boats and the percent observer coverage ranges from 10% to 100%, depending on the fishery. Indices of the relative mortality imposed by fisheries discards and bycatch were calculated here as the ratio of the observed numbers of crab discarded or bycaught in a given year divided by the estimated population numbers in a given year. Only discard mortality is considered for the directed fishery in our models because the range of sizes modeled exclude the largest males, which are the targets of the commercial fishery for snow crab.

### Crab density
The numbers of crab estimated from the population dynamics models were also used as covariates in the GAMs. Increasing densities of crab could capture aspects of intraspecific competition not captured in other covariates. Each respective model of mortality incorporates the population size of the corresponding maturity state given their spatial co-occurence. Immature mortality also incorporates mature population size because crab are thought to move more extensively after maturing in the pursuit of mates, which suggests that their overlap with the immature portion of the population could be larger than the snapshot the survey provides. This increased overlap could result in impacts on mortality, hence the inclusion of mature population size in the immature mortality models.

## Generalized additive models
Generalized additive models (GAMs) were used in the R programming language (package mgcv; Wood, 2011) to relate changes in estimated mortality by maturity state and year, $m_{p,y}$ to environmental covariates by maturity state and year, $\phi_{m,y}$, because of their flexibility in fitting potential non-linear relationships. Models were first fitted in which all relevant covariates were included in the model of the form:

@@

Caloric requirements increased sharply in 2018 and to explore potential impacts of this increase, we analyzed the weight at size data available (\autoref{wt_size_year_temp}). GAMs were used to predict observed weights at size $w_{i,s,y}$ using the bottom temperature in which the crab was collected, $t_{i}$, measured carapace width $cw_{i}$, and year as a factor:

@@

include_graphics("plots/BSFRF_map_plots.jpg")

>>>>

---
author: ""
date:
output:
  pdf_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 6
    keep_tex: true
    includes:
      in_header: my_header.tex
header-includes: 
  - \usepackage[left]{lineno}
  - \linenumbers
#-  \pagenumbering{gobble}
csl: fish-and-fisheries.csl
toc: yes
title: "Supplementary materials for 'Canaries of the Arctic: the collapse of eastern Bering Sea snow crab'"
---

@@

# Supplementary materials
## Methods overview
We used an integrated population model to estimate variation in mortality over time for snow crab in the eastern Bering Sea and generalized additive models (GAMs) to relate the estimated variation in mortality to potential stressors in the environment.  The population dynamics model was fit to abundance and size composition data from the National Marine Fisheries Service (NMFS) summer bottom trawl survey on the eastern Bering Sea shelf to estimate total mortality by maturity state and year for male snow crab. We then developed indices for temperature occupied, disease prevalence, cannibalism, and crab density from the NMFS survey to test as covariates in GAMs. Cod predation indices were developed using stomach content data collected on NMFS surveys in addition to cod size composition and abundances. Indices for fishery related effects were collated from fisheries statistics from the Alaska Department of Fish and Game and also included in the GAMs.

Ecological detective work in the marine environment is hampered by the difficulty of observation and this is particularly so on the eastern Bering Sea shelf. The waters in which snow crab reside range from 50-200 meters deep and are seasonally covered by ice, making data collection only feasible in the summer. Consequently, the survey based portions of our analyses are derived from a yearly snapshot of the population over an approximately 30 year period. Each of the hypotheses explored here clearly result in some mortality. We know that millions of crab are eaten by cod every year, the directed and bycatch fisheries kill crab, larger crab eat smaller crab, and crab die from bitter crab disease each year. The goal of our analysis is to place each of these processes in a historical context to try to understand the relative impact of each and what was different about the recent collapse. More than one way exists to analyse the available data on this issue. Below we describe our approach, including a description of each of the components of our analysis, a discussion of the rationale behind our modeling decisions, and sensitivities and simulation tests of our models, all of which provide what we think is sound reasoning for our analysis.

## Population dynamics model
The population dynamics model presented here incorporated the best available information on relevant population processes to estimate total mortality for male snow crab on the eastern Bering Sea shelf and is similar in structure to the model used to assess eastern Bering Sea snow crab for management (Szuwalski, 2021). The model tracked numbers of male crab at size at maturity state over time with size bins ranging from 30-95 mm carapace width with 5 mm bin widths. Only male crab were modeled because male and female crab appear to have somewhat different dynamics and the male crab in the modeled size range are better selected by the survey gear (Szuwalski, 2021). Snow crab are sexually dimorphic, with male snow crab growing to nearly twice the size of females, which accounts for the better selection in the survey. Only crab smaller than 95 mm were modeled for two reasons: 1) to attempt to isolate the effect of the directed fishery (crabs of >101 mm carapace width are targeted in the fishery; discussed further below) and 2) almost all of the crab that disappeared since 2018 are in this size range. The population dynamics model operates on a half year time step, starting in July at the time of the NMFS survey.  Total mortality (Z) is estimated by year (y) and maturity state (m). Other estimated parameters include the initial numbers at size by maturity state, yearly log recruitments, a vector of scalars that determine the proportions of estimated recruitment split into the first two size bins, and a variance component for the penalty on total mortality. Parameters determining growth, maturity, and survey selectivity were estimated outside of the model and specified when estimating mortality and catchability. Mortality is the only population process that occurs in the first half of a given year:

@@

The size transition matrix $X_{s,s'}$ was constructed using growth increment data collected over several years (see Szuwalski [2021] for a summary) to estimate a linear relationship between pre- and post-molt carapace width (\autoref{linear_growth}), ($\hat{W}^{pre}_{s,w}$ and $\hat{W}^{post}_{s,w}$, respectively) and the variability around that relationship was characterized by a discretized and renormalized normal distribution with a size-varying standard deviation, Y~*s,w,w'*~ (\autoref{linear_growth}).

@@

Finally, the last half of the year of mortality is applied to the population after growth, molting, and recruitment occurs. Note that this allows a crab to experience two different mortalities within a given year as it undergoes terminal molt.

@@

The observed numbers of crab at size by year in the NMFS survey reflect the ability of the trawl gear to capture the crab, also known as 'selectivity'. The selectivity of trawl gear can change according to size, and consequently needs to be accounted for in the population dynamics model when fitting to the survey data. Values for survey selectivity at size were specified using data from experimental *Nephrops* trawls (a small trawl net designed to maintain bottom contact), operated by the Bering Sea Fisheries Research Foundation in collaboration with the NMFS summer survey. The experimental trawls were performed at the same time and location as the NMFS summer survey tows to evaluate the efficiency of the NMFS survey trawl gear at capturing snow crab (Somerton et al., 2013). The *Nephrops* gear used by the BSFRF was assumed to capture all crab in its path given strong bottom contact.  The resulting area-swept estimates of numbers of crab at size from the BSFRF and NMFS surveys ($\hat{N}_{y,s,NMFS}$ and $\hat{N}_{y,s,NMFS}$, respectively) can be used to infer the selectivity of the NMFS gear in year y as:

 \begin{equation} S_{y,NMFS} = \frac {\hat{N}_{y,s,NMFS}}{\hat{N}_{y,s,BSFRF}} \end{equation}

@@

### Objective function
The objective function for the population dynamics model consists of likelihood components (representing the fit of the model to the data) and penalty components (which incorporate constraints in the fitting based on prior information) that are summed and minimized in log space to estimate parameters within the model. Several data sources were fit to using the following likelihoods. Observed size composition data for immature and mature males were fit using multinomial likelihoods and were implemented in the form:

@@

L~*x*~ was the likelihood associated with data component x, where $\lambda_{x}$ represented an optional additional weighting factor for the likelihood, $N_{x,y}$ was the sample sizes for the likelihood, $p^{obs}_{x,y,l}$ was the observed proportion in size bin *l* during year *y* for data component *x*, and $\hat {p}_{x,y,l}$ was the predicted proportion in size bin *l* during year *y* for data component *x*. Sample sizes were input as 50.

@@

Smoothing penalties were placed on estimated vectors of deviations for immature and mature natural mortality (and immature and mature catchability in the simulation analyses aimed at understanding the estimability of mortality and catchability) using normal likelihoods on the second differences of the vectors. Normal priors were also placed on the mean value of natural mortality and catchability and the deviation of the estimated mortality from that mean. A prior value of 0.27 is used for the average natural mortality based on assumed maximum age of 20 and Hamel's (2015) empirical analysis of life history correlates with natural mortality. The priors used for catchability were derived from the selectivity experiments described above.  The normal priors were of the form:

\begin{equation} P_{x} = \lambda_{x} \sum_{y} \frac {((\hat{I}_{x,y}) - (I_{x,y}))^2}{CV_{x,y}^2}
\end{equation}

$P_{x}$ was the contribution to the objective function of the penalty associated with model estimate *x*, $\lambda_{x}$ was any additional weighting applied to the component, $\hat{I}_{x,y}$ was the predicted value of population process *I* relevant to penalty *x* during year *y*, I~*x,y*~ was the prior value of process *I* relevant to penalty *x* during year *y* and CV~*x,y*~ was the input coefficient of variation for penalty *x* during year *y*.

An example of the way in which these equations were implemented can be seen in lines 132-218 of 'snow_down.TPL' in our github repo 'snow_down/models/model_vary_m'.

## Population dynamics model sensitivities
Modeling decisions are necessarily made in the process of writing population dynamics models and it is possible for these decisions to influence the outcome of an analysis. Within the context of our model, these decisions include what processes to allow to vary over time, the weights assigned to different data sources and penalties in the objective function, which parameters to place priors or penalties on, and what those priors or penalties should be. We ran several sensitivity analyses to understand the implications of these modeling decisions on the outcome of our analysis.

### Does allowing mortality or catchability to vary over time improve model fits?
Catchability and mortality are somewhat confounded within population dynamics models (Thompson, 1994). Fewer crab observed in a given year can be attributed to either crab dying or by crab moving out of the surveyed area either by walking out of the boundaries or burying themselves into the substrate. At the same time, it is also clear that catchability and mortality likely vary over time in reality in spite of the fact that they are often assumed to be time-invariant in population dynamics models (Johnson et al., 2014). Somerton et al. (2013) showed that catchability varied somewhat by substrate and depth for snow crab in the EBS. The spatial distribution of snow crab varies over time and substrate and depth vary over space, so it follows that catchability should also vary over time.

We started exploring the impacts on model output of including time-variation in mortality and catchability by fitting a model with no time-variation in mortality or catchability. Then we compared the output of this model to models that allow time-variation in mortality, catchability, and both processes simultaneously (\autoref{inc_complex} & \autoref{inc_comp_proc}). The model with no time-variation in mortality or catchability was able to capture the general trend in immature and mature survey abundance solely through estimating variability in recruitment. Allowing time-variation in catchability improved the fits to immature survey abundances more than time-varying mortality, but time-variation in either process improved fits in a similar manner for mature survey abundances. Mature size composition data were fit similarly for all models, but immature size composition data were better fit by the models that allowed time-varying catchability (\autoref{inc_complex}). Part of the reason this difference in fits to immature size composition data occurs is the variability in the first several size bins resulting from the poor selectivity of the survey for small animals. Sometimes the peaks seen in larger size classes are reflected in the preceding years' data for the smallest size classes, sometimes those peaks are not reflected (compare \autoref{expand_size_fits_imm} to \autoref{expand_size_fits_mat}). As a consequence, positive residuals occur in the smallest size classes when a pseudocohort is consistently seen in large size classes, but not observed in the smallest size bins (e.g. 1991 vs. 1992; 1997 vs 1998).

@@

Differences among sensitivity scenarios resulted in very small changes in the fits to the data (\autoref{sens_fits}), but larger changes in estimated mortalities and catchabilities (\autoref{sens_ests}). The smoothness penalty placed on mortality over time appeared to be the largest driver of changes in estimates of M and q, so we looked at a wider range of smoothness penalties (i.e. 0.001, 0.1, 0.25, 0.5, 1, 5, 10, 1000).  Trajectories of mortalities were roughly preserved across this range. The prior on mean natural mortality predictably scaled the estimated time series up or down. The best available information suggests natural mortality should be approximately 0.27 given an assumed (but based on a range of studies; see Szuwalski, 2021 for a summary) maximum age of 20 years for wild snow crab. Based on these analyses, we elected to use small smoothing penalties because there is no evidence to suggest that mortality should be particularly smooth from year to year and relatively tight priors on the mean mortality given outside information to support an average mortality value based on longevity. These analyses also underscore the fact that the scale of the population is difficult to estimate with the data available and the need to relate mortality to the environmental covariates outside of the population dynamics model. This likely comes from the fact that recruitment and immature mortality are confounded (i.e. fewer immature crab in a given year can be because of increased immature mortality or because of lower recruitment).

@@

Another important point for consideration in covariate construction is the estimation of mortality by maturity state. Snow crab in the EBS undergo an ontogenetic migration in which juvenile crab settle on the northeast portion of the shelf after their pelagic phase, then migrate southwest into deeper and (usually) warmer waters (Ernst et al., 2005; Parada et al., 2010). This means that the conditions and stressors experienced by immature crab can be different than those by mature crab. To address this issue, the spatial data sets for temperature, disease, and cannibalism were split based on the size above which half of the population was mature in a given year. The size at which more than half of the population is mature changes by year, depending on recruitment dynamics and other demographic processes (\autoref{split_mat}). After the survey data were split at the 50% at maturity size, time series of maturity-specific environmental stressors (\autoref{grid_covar}) were created as described below.

### Temperature
Temperature is one of the key physical variables that structures the benthic ecosystem of the EBS (Mueter and Litzow, 2008). The cold pool, a mass of water <2 degrees Celsius, can act as a barrier to species interaction based on temperature preferences of different species. Snow crab are a stenothermic species, preferring cold water and juvenile snow crab in particular are rarely found outside of the cold pool (Dionne, 2003). The cold pool is directly related to the winter ice extent in the Bering Sea and has varied dramatically over time as the ecosystem moves between cool and warm stanzas (e.g. 2006-2010 vs. 2014-2019; Figure 1b of the main text and \autoref{bottom_temp}). As the cold pool changes from year to year, so does the spatial distribution of snow crab (\autoref{snow_dist}). The ontogenetic migration of snow crab results in crab of different sizes and maturity states experiencing different temperatures in a given year (\autoref{temp_occupied}). The 'temperature occupied' for different sizes of crab by year $T_{s,y}$ was calculated here as an average of the observed bottom temperatures at the stations at which crab of a given size were captured $t_{i}$, weighted by the area-swept density of crab at a given station $d_{i}$:

@@

### Predation
Pacific cod (*Gadus macrocephalus*) are the most important predator of snow crab based on stomach content data collected in the NMFS bottom trawl survey (Long and Livingston, 1998), with 16.5% of cod stomachs containing snow crab (Burgos et al., 2010). Crab ranging from 8-57 mm carapace width constitute 95% of the crab consumed by cod in the Bering Sea, but crab up to 106 mm carapace have been observed in cod stomachs (Burgos et al., 2010). An index of summer daily consumption (tons/day) of snow crab between 30-95mm carapace width eaten by Pacific cod in the eastern Bering Sea was developed using cod stomach content data from the survey to estimate the proportion by weight of crab in cod diets and the size composition of crab by carapace width of prey found in cod stomachs, stratified by year, survey stratum, and cod length (collection and analysis methods described in Livingston et al. 2017). Cod total consumption rate (metabolic demand) was calculated using a cod bioenergetics model (Holsman and Aydin 2015) to estimate laboratory-measured maximum consumption rates adjusted for bottom water temperatures and cod abundance-at-length measured at each haul location (following methods described in Barbeaux et al. 2020), and summed to an eastern Bering Sea ecosystem-wide total.

Changes in the cold pool can alter the interaction between snow crab and Pacific cod over time. Decreases in the size of the cold pool coincide with more northerly positions of the centroids of abundance of cod (e.g. 2003 and 2018-2019; \autoref{cod_centroids} & \autoref{cod_pred_data}). This increased interaction coincided with increased numbers of crab consumed by cod in the last several years (\autoref{cod_consumption_raw}). The estimated number of cod greater than 50 cm was also near all-time highs around the period during which crab collapsed (\autoref{cod_sensitivity}). However, this period of time also coincided with the appearance of the largest pseudo-cohort of snow crab ever seen in the Bering Sea. Given the generalist nature of Pacific cod, one would expect to see an increase in the amount of crab consumed by cod during this period of time even if there weren't differences in the interactions between the species as a result of changes in the cold pool or increases in abundance of large cod.  To evaluate the possibility cod consumption has influenced the mortality of snow crab over time, the relative impact of consumption with respect to the population size must be considered. Predation indices were calculated for crab by year $P_{m,y}$ by calculating the ratio of the extrapolated biomass of crab consumed by cod to the estimated biomass of crab, $N_{y,m,s}*w_{s}$:

@@

The exact amount of crab eaten cannot be calculated from the available diet data because they are a snapshot of consumption at one point during the year and consumption would be expected to change with spatial overlap and temperature-driven changes in metabolism occurring throughout the year. Consequently, removals due to predation cannot be directly incorporated into the model as fishery removals might be. 
The index of consumption described above incorporates the most available data on cod predation, but some strong assumptions are made (e.g. summer diet is representative of the entire year). As a sensitivity to these assumptions, we also tested the ratio of the number of cod greater than 50 cm to crab abundance in a given year as an alternative index of predation in the GAMs. Ultimately, changing the index of predation did not impact the results of the fitting of the GAMs; temperature and mature population size were still the only significant covariates and the estimated shapes of relationships and deviance explained were very similar between models with the different predation indices. Consequently, the models presented in the main text use the index of consumption as the predation index because it uses the most available information on cod predation (i.e. stomach contents and the abundance and size composition of cod).

@@

The proportion of smaller than 55 mm carapace width crab overlapping with larger than 95 mm carapace width crab was calculated by finding the intersection of the station IDs at which at least one crab of both size classes was observed. The density of crab larger than 95 mm carapace width was calculated as the number of >95 mm carapace width crab observed at those stations multiplied by the area swept. This exercise was also done by 5 mm size bins to show the overlap of small crab of different sizes with large crab (\autoref{cannibalism_size_raw}). The final index aggregated all crab smaller than 55 mm carapace width (\autoref{cannibalism}). Indices of cannibalism were only included in the immature models given laboratory observations indicate cannibalism is rare among crab of similar sizes, though molting crab can be vulnerable.

### Fisheries data
Snow crab are caught both in a directed fishery (i.e. a fishery aimed at capturing snow crab) and non-directed fisheries (i.e. fisheries with targets other than snow crab). In the directed fishery, under-sized and/or dirty shelled male crab are often discarded and all females are discarded. Snow crab are discarded from non-directed fisheries using a variety of gear types (including trawl, pots, hook-and-line) and targeting a variety of species (e.g. Pacific cod, walleye pollock, and yellowfin sole) that operate over a wide fraction of the Bering Sea shelf (\autoref{bycatch_binned}). \autoref{bycatch_binned} is plotted in log space, so it appears that the bycatch is spread widely over the shelf, but in normal space, the bycatch is more concentrated (e.g. \autoref{bycatch_binned_2018}). The location of the centroids of the bycatch have moved over time and increases in latitude correspond with warm years in which reduced ice extent allowed for fishing farther north (\autoref{bycatch_centroids}). Bycatch in  trawl fisheries are by far the largest sources of bycatch mortality (\autoref{bycatch}). Data on discards and bycatch of snow crab are collected by at-sea observers on fishing boats and the percent observer coverage ranges from 10% to 100%, depending on the fishery. Some fraction of the mortality imposed by non-directed fleets is likely unobserved due to crab being struck by the gear and not captured. Consequently, indices of the relative mortality imposed by fisheries discards and bycatch were calculated here as the ratio of the observed numbers of crab discarded or bycaught in a given year divided by the estimated population numbers in a given year. Only discard mortality is considered for the directed fishery in our models because the range of sizes modeled exclude the largest males, which are the targets of the commercial fishery for snow crab.

### Crab density
The numbers of crab estimated from the population dynamics models were also used as covariates in the GAMs. Changing densities of crab could capture aspects of intraspecific competition not captured in other covariates. Each respective model of mortality incorporates the population size of the corresponding maturity state given their spatial co-occurence. Immature mortality also incorporates mature population size because crab are thought to move more extensively after maturing in the pursuit of mates, which suggests that their overlap with the immature portion of the population could be larger than the snapshot the survey provides. This increased overlap could result in impacts on mortality, hence the inclusion of mature population size in the immature mortality models.

## Generalized additive models
Generalized additive models (GAMs) were used in the R programming language (package mgcv; Wood, 2011) to relate changes in estimated mortality by maturity state and year, $m_{m,y}$ to environmental covariates by maturity state and year, $\phi_{m,y}$, because of their flexibility in fitting potential non-linear relationships. Models were first fitted in which all potential relevant covariates were included in the model of the form:

@@

Caloric requirements increased sharply in 2018 and to explore potential impacts of this increase, we analyzed the weight at size data available (\autoref{wt_size_year_temp}). A GAM was used to predict observed weights at size $w_{i,s,y}$ using the bottom temperature in which the crab was collected, $t_{i}$, measured carapace width $cw_{i}$, and year as a factor:

@@

include_graphics("plots/BSFRF_map_plots.png")
